Sampling Strategies in Decisions from Experience

### 2022-05-26 manuscript.Rmd
Update abstract

abstract: |
  Add abstract

keywords          : "risky choice, information sampling, evidence accumulation, computational modeling, cumulative prospect theory"
wordcount         : ""

@@

The aim of this paper was to sketch how the mind might exercise the capacity of making decisions from experience (DFE).
In DFE, people cannot rely on symbolic descriptions of the choice alternatives, but only on samples of past experiences.
One paradigm used to study DFE experimentally is the sampling paradigm, where people can---as much as they want, in the way they want, and without costs---sample the possible outcomes of prospects before making a final choice between them.
To approach an explanation for how people proceed in such a paradigm, this paper presented two models that integrate the idea of a potential interplay between sampling and decision strategies into the computational framework of sequential sampling models.

@@

Hills and Hertwig [-@hillsInformationSearchDecisions2010] proposed that the way people sample the outcomes of prospects (sampling strategies) may be linked to the way they process the outcomes to evaluate the prospects and make a corresponding choice (decision strategies).
Specifically, the authors assumed two potential links between what they considered paradigmatic instances of sampling and decision strategies:
People either switch prospects after each sampled outcome (piecewise sampling) to compare single outcomes over repeated rounds (round-wise decisions), or they switch only once between prospects (comprehensive sampling) and make only one comparison between the means across all sampled outcomes (summary decisions).
The authors provided tentative empirical support for their claim by showing that a larger proportion of the choices of people who, on average, switched infrequently between prospects were better predicted by a summary strategy (rather than a round-wise strategy), and vice versa.
Moreover, the authors discussed and demonstrated in a simulation based on empirical sampling data that round-wise decisions may cause an as-if underweighting of rare outcomes pattern in DFE and thus, contribute to the decision-experience gap.

@@

2. The sequential sampling framework was used as a modeling outlet to incorporate the idea of a potential interplay between sampling and decision strategies into formal models that offer a complete account of the decision-making process in the sampling paradigm and make theoretical commitments explicit.

3. Cumulative prospect theory was used to model simulated choice data, allowing to inquire whether the choices implied by the potential interplay between sampling and decision strategies may translate to characteristic signatures in CPT’s psychoeconomic value and weighting function.

@@

Relying on the sequential sampling framework, the decision-making processes were modeled as evidence accumulation over time. 
Two different models were presented ...

In a computational analysis, the presented model was used to simulate how the proposed sampling and accumulation processes shape the final choice.
Therefore, the switching probabilities and evidence thresholds were systematically varied, while simulating choices between a safe and a risky prospect.
The model predicts the choices to vary systematically with the switching probability.
Specifically, whereas the choice patterns correspond closely to the solutions of EV maximization for low switching probabilities, increases in the switching probability can lead to an increasingly pronounced as-if underweighting of rare outcomes pattern.
This is because for high switching probabilities, each independent mean comparison is based on small samples, in which small-probability outcomes are underrepresented rather than overrepresented.
As a consequence, then, small-probability outcomes are predicted to contribute to the accumulated evidence less rather than more than would be warranted by their objective probabilities.
Accordingly, the analyses showed that the structure of the choice problem---i.e., whether a small-probability outcome is present and which value it takes---is an important moderator for the effect of sampling strategies on the choice pattern. 
Importantly, while this explanation emphasizes that the reliance on small samples is a strong contributor to the as-if underweighting of rare outcomes pattern in DFE, it has little to do with the common notion of sampling error [e.g., @foxDecisionsExperienceSampling2006; @hertwigDecisionsExperienceEffect2004].
That is, the as-if underweighting of rare outcomes is not caused by a deviation of the sampled relative frequencies from the objective probabilities, but by the way the sampled outcomes are integrated and processed to evaluate the prospects.
The analyses made this clear by showing that an as-if underweighting of rare outcomes pattern occurred, even if the sampling error is controlled for.
In sum, by explicitly modeling their link, the current computational analysis showed how the interplay of sampling and decision strategies can shape the final choice.

The current paper further extended the analysis of @hillsInformationSearchDecisions2010 by showing that the effects of sampling and decision strategies on the final choice can translate to characteristic signatures in cumulative prospect theory's value and weighting function. 
Having controlled for sampling error, the estimated CPT parameters and the resulting shapes of the value and weighting function primarily reflected how sensitive the combinations of sampling and decision strategies are to particular changes in the outcomes and their probabilities.
Thereby, the analyses extends the notion that CPT can be used as a measurement tool to characterize information processing [@pachurHowTwainCan2017] from risky to uncertain choice.

@@

### Why Sampling and Decision Strategies?

>>>>

abstract: |
  In the sampling paradigm, people can---as much as they want, in the way they want, and without costs---sample the possible outcomes of prospects before making a final choice. Do the sequential pattern in which the outcomes are sampled (sampling strategy) and the way the sampled outcomes are integrated and processed to evaluate the prospects (integration/decision strategy) act together to shape the choice? This paper presents two models that incorporate the potential interplay between both types of strategies into the sequential sampling framework. Each model assumes an evidence accumulation process representing a distinct integration/decision strategy and is used to simulate choices between a safe and a risky prospect, while systematically varying sampling strategies. The simulations show that the potential interplay between both types of strategies can lead to systematically different choice patterns and can contribute to the as-if underweighting of rare outcomes pattern in decisions from experience, irrespective of the existence or absence of sampling error. The model-implied differences in the choice patterns translate to distinct signatures in the psychoeconomic functions of cumulative prospect theory.

keywords          : "decisions from experience, information sampling, evidence accumulation, computational modeling, cumulative prospect theory"
wordcount         : ""

@@

The aim of this paper was to sketch how the mind might exercise the capacity of making decisions from experience (DFE).
In DFE, people cannot rely on symbolic descriptions of the choice alternatives, but only on samples of past experiences.
One paradigm used to study DFE experimentally is the sampling paradigm, where people can---as much as they want, in the way they want, and without costs---sample the possible outcomes of prospects before making a final choice.
To approach an explanation for how people proceed and choose in such a paradigm, this paper presented two models that integrate the potential interplay between sampling and integration/decision strategies into the sequential sampling framework.

@@

Hills and Hertwig [-@hillsInformationSearchDecisions2010] proposed that the way people sample the outcomes of prospects (sampling strategy) may be linked to the way they process the outcomes in order to evaluate the prospects and make a corresponding choice (integration/decision strategy).
Specifically, the authors assumed two potential links between what they considered paradigmatic instances of both types of strategies:
People either switch prospects after each sampled outcome (piecewise sampling) to compare single outcomes over repeated rounds (round-wise integration/decision), or they switch only once between prospects (comprehensive sampling) and make one comparison between the means across all sampled outcomes (summary decision).
The authors provided tentative empirical support for their claim by showing that a larger proportion of the choices of people who, on average, switched infrequently between prospects were better predicted by a summary strategy (rather than a round-wise strategy), and vice versa.
Moreover, the authors discussed and demonstrated in a simulation based on empirical sampling data that round-wise decisions may cause an as-if underweighting of rare outcomes pattern in DFE and thus, contribute to the decision-experience gap.

@@

2. The sequential sampling framework was used as a modeling outlet to incorporate the potential interplay between sampling and decision strategies into formal models that offer a complete account of the decision-making process and make theoretical commitments explicit.

3. Cumulative prospect theory was used to model simulated choice data, allowing to inquire whether the choices implied by the potential interplay between both types of strategies may translate to characteristic signatures in CPT’s psychoeconomic value and weighting function.

@@

This paper started from the computational problem the mind is trying to solve in the sampling paradigm, and the principle logic that governs its solution:
In order to make systematic choices between prospects, the mind must assess true differences in their latent properties, and it can only do so by searching for evidence in information samples.
Relying on the sequential sampling framework, the current paper assumed that this search process takes the form of an evidence accumulation over time, where outcomes are sampled until an evidence threshold is reached.
Two different models were presented, each of which assumed an accumulation process that resembles a distinct decision strategy.
The round-wise integration model assumed that the means across sets of sampled outcomes are repeatedly compared and a choice is made once the number of won comparisons reaches a threshold.
The summary integration model assumed that all outcomes sampled from a prospect are integrated into the same cumulative sum and that a choice is made once the sum reaches a threshold.

The current paper discussed and demonstrated in a computational analysis how sampling strategies may or may not alter the latent property for which each of the models accumulates evidence, and thus may or may not cause each model to produce systematically different choice patterns:
Specifically, in the round-wise integration model, changes in the sampling strategy alter the degree to which evidence is accumulated for differences in either the expected values or in the probability to produce a higher outcome most of the time.
If the prospect that produces the higher outcome most of the time does not also possess the higher expected value, changes in the sampling strategy come to imply systematically different choice patterns.
The simulations showed that high switching probabilities led the round-wise integration model to produce systematic deviations from EV maximization that take the form of an as-if underweighting of rare outcomes, even if sampling error is controlled for. 
In contrast, in the summary integration model, changes in the sampling strategy do not cause  alter which latent property evidence is accumulated for, but rather

Finally, it was shown how the effects of the potential interplay between sampling and integration/decision strategies on the final choice can translate to characteristic signatures in cumulative prospect theory's value and weighting function. 
Having controlled for sampling error, the estimated CPT parameters and the resulting shapes of the value and weighting function primarily reflected how sensitive the combinations of sampling and decision strategies are to particular changes in the outcomes and their probabilities.
Thereby, the analyses extends the notion that CPT can be used as a measurement tool to characterize information processing [@pachurHowTwainCan2017] from risky to uncertain choice.

@@

## Why Particular Sampling and Decision Strategies Could Be Linked?
